Concurrent Programming in Harmony

### 2021-12-24 book/paper.tex
Finished pass over Chapter 10

The code in \autoref{fig:lockspec} is similar to the code in
Harmony's \texttt{synch} module.
(The module generalizes locks to \emph{binary semaphores}
(\autoref{ch:sbs}), but the lock interface is the same.)
\autoref{fig:cssynch} shows how locks may be used to implement
a critical section.
\autoref{fig:incfixed} gives an example of how locks may be used
to fix the program of \autoref{fig:inc}.

@@

In \autoref{fig:tas} we have shown a lock implementation based on a shared
variable and a private variable for each thread.   The private
variables themselves are actually implemented as shared variables,
but they are accessed only by their respective threads.
A thread usually does not keep explicit track of whether it has a lock
or not, because it is implied by the control flow of the program---a
thread implicitly \emph{knows} that when it is executing in a critical
section it has the lock.
There is no need to keep \textit{private} as a shared
variable---we only did so to be able to show and check the invariants.
\autoref{fig:tas2} shows a more straightforward implementation of a spinlock.
% The solution is similar to the na\"{i}ve solution of \autoref{fig:naivelock},
% but uses test-and-set to check and set the lock variable atomically.
The lock is also cleared in an atomic statement to prevent a data race.
This approach is general for any number of threads.

You can test the spinlock with the program in \autoref{fig:cssynch}
using the command \texttt{harmony -m synch=taslock code/cssynch.hny}.
The \texttt{-m} flag tells Harmony to use the \texttt{taslock.hny} file
rather than the standard \texttt{synch} module (which only contains
a specification of the lock methods).

\index{starvation}%
The spinlock implementation suffers potentially from \emph{starvation}:
an unlucky thread may never be able to get the lock while other threads
successfully acquire the lock one after another.  It could even happen
with just two threads: one thread might successfully acquire the lock
repeatedly in a loop, while another thread is never lucky enough the
acquire the thread in between.
% \autoref{ch:starvation} will talk more about starvation and how
% to prevent it.

A \emph{ticket lock} (\autoref{fig:ticket} is an implementation of a
lock that prevents starvation using an atomic \emph{fetch-and-increment}
operator.  It is inspired by European bakeries.  A European bakery has a
clearly displayed counter (usually just two digits) and a ticket dispenser.
Tickets are numbered 0 through 99 and repeat over and over again
(in the case of a two digit counter).
When a customer walks into the bakery, they draw a number from the dispenser
and wait until their number comes up.
Every time a customer has been helped, the counter is incremented.
(Note that this only works if there can be no more than 100 customers in
the bakery at a time.)

\autoref{fig:ticket} similarly uses two variables for a lock,
\textit{counter} and \textit{dispenser}.  When a thread acquires the
lock, it fetches the current dispenser value and increments it
modulo \texttt{MAX\_THREADS}, all in one atomic operation.
In practice, \texttt{MAX\_THREADS} would
be a number like $2^{32}$ or $2^{64}$, but since the Harmony model checker
checks every possible state, limiting \texttt{MAX\_THREADS}
to a small number such as done
here will significantly reduce the time to model check a Harmony program.
Plus it is easier to check that it fails when you run it with more than
\texttt{MAX\_THREADS} threads.
You can test the implementation using the command
\texttt{harmony -m synch=ticket code/cssynch.hny}.
To see it fail, try
\texttt{harmony -c NTHREADS=10 -m synch=ticket code/cssynch.hny}.

@@

In most operating systems, threads are virtual (as opposed to ``raw CPU cores'')
and can be suspended until some condition changes.
For example, a thread that is trying to acquire a lock can be suspended until the lock is
available.
In Harmony, a thread can suspend itself and save its context (state) in a
shared variable.  Recall that the context of a thread contains
its program counter, stack, and registers.
A context is a regular (if complex) Harmony value.
The syntax of the expression that a thread executes to suspend itself
is as follows:

@@

You will find that using the \emph{implementation} of a lock instead
of the \emph{specification} of a lock (in the \texttt{synch} module)
often leads to the model checker searching a
significantly larger state space.
Thus it makes sense to model check larger programs in a modular fashion:
model check one module implementation at a time, using
specifications for the other modules.

\section*{Exercises}
\begin{problems}
\item
Run \autoref{fig:incfixed} using (i) \texttt{synch} and (ii) \texttt{synchS}.
Report how many states were explored by Harmony for each module.
\item \label{ex:xy} \autoref{fig:xy} shows a Harmony program with two variables <{x}>
(initially 0) and <{y}> (initially 100) that can be accessed through methods
\texttt{setX} and \texttt{getXY}.  An application invariant is that \texttt{getXY}
should return a pair that sums to 100.  Add the necessary synchronization code.
\item \label{ex:trylock} Implement \texttt{tryAcquire}(\textit{b}) as an additional
interface for both the \texttt{synch} and \texttt{synchS} modules.
This interface is like \texttt{acquire}(\textit{b}) but never blocks.  It
returns <{True}> if the binary semaphore was available (and now acquired) or <{False}>
if the binary semaphore was already acquired.
Hint: you do not have to change the existing code.
\item People who use an ATM often first check their balance and then withdraw
a certain amount of money not exceeding their balance.  A negative balance
is not allowed.  \autoref{fig:atm} shows two operations on bank accounts:
one to check the balance and one to withdraw money.
Note that all operations on accounts are carefully protected by a lock
(i.e., there are no data races).
The \texttt{customer}
method models going to a particular ATM and withdrawing money not exceeding
the balance.
Running the code through Harmony reveals that there is a bug.
It is a common type of concurrency bug known as \emph{Time Of Check Time Of
Execution} (TOCTOE).
\index{Time Of Check Time Of Execution}
\index{TOCTOE}
In this case, by the time the withdraw operation is performed,
the balance can have changed.
Fix the code in \autoref{fig:atm}.  Note, you should leave the
customer code the same. You are only allowed to change the
\texttt{atm\_} methods, and you cannot use the \textbf{atomically}
keyword.
\end{problems}

@@

\begin{figure}
\begin{code}
<{:input:../code/mesa.hny}>
\end{code}
\caption{\harmonylink{modules/synch.hny}
Implementation of condition variables in the \texttt{synch} module}
\label{fig:cv}
\end{figure}

@@

Harmony provides support for Mesa monitors
through the Harmony \texttt{synch} module.
\autoref{fig:cv} shows the implementation of
condition variables in the \texttt{synch} module.
\texttt{Condition}() creates a new condition variable.
It is represented by a dictionary containing
a bag of contexts of threads waiting on the condition variable.
(The \texttt{synchS} library instead uses a list of contexts.)

@@

This completes all the Havender methods.
There is, however, another approach, which is sometimes called deadlock
\emph{avoidance}
\index{deadlock avoidance}%
instead of deadlock \emph{prevention}.
In the case of the Dining Philosophers, we want to avoid the situation where each
diner picks up a fork.  If we can prevent more than four diners from starting to
eat at the same time, then we can avoid the conditions for deadlock from ever
happening.
\autoref{fig:dinersavoid} demonstrates this concept.  It uses a
\emph{counting semaphore} to restrict the number of diners at any time to
four.  A counting semaphore is like a binary semaphore, but can be
acquired a given number of times.  It is supported by the \texttt{synch}
module.  The \texttt{P} or ``procure'' operation acquires a counting
semaphore.  That is, it tries to decrement the semaphore, blocking while
the semaphore has a value of 0.  The \texttt{V} or ``vacate'' operation
increments the semaphore.

@@

The \texttt{synch} library supports a synchronized message queue,
similar to the \texttt{Queue} object in Python.
Its interface is as follows:
\begin{itemize}
\item \texttt{Queue()} returns a new message queue;
\item <{put}>(<{q}>, \textit{item}) adds \textit{item} to the queue pointed to by <{q}>;
\item <{get}>(<{q}>) waits for and returns an item on the queue pointed to by <{q}>.
\end{itemize}

@@

It is useful to create an abstract network that reliably sends messages
between threads, much like the FIFO queue in the \texttt{synch} module.
For this, we need a network protocol that Alice and Bob can run.
In particular, it has to be the case that if Alice sends application
messages $m_1, ..., m_n$ in that order, then if Bob receives an
application message
<{m}>, then $m = m_i$ for some <{i}> and Bob will already have received
application messages $m_1, ..., m_i$ (safety).
Also, if the network is fair and Alice sends application message
<{m}>, then eventually Bob should deliver <{m}> (liveness).

@@

\section{The \texttt{synch} module}

\index{synch module}%
The \texttt{synch} module provides the following methods:

>>>>

The code in \autoref{fig:lockspec} is similar to the code in
Harmony's <{synch}> module.
(The module generalizes locks to \emph{binary semaphores}
(\autoref{ch:sbs}), but the lock interface is the same.)
\autoref{fig:cssynch} shows how locks may be used to implement
a critical section.
\autoref{fig:incfixed} gives an example of how locks may be used
to fix the program of \autoref{fig:inc}.

@@

\autoref{fig:tas} shows a lock implementation based on a shared
variable and a private variable for each thread.   The private
variables themselves are actually implemented as shared variables,
but they are accessed only by their respective threads.
A thread usually does not keep explicit track of whether it has a lock
or not, because it is implied by the control flow of the program---a
thread implicitly \emph{knows} that when it is executing in a critical
section it has the lock.
There is no need to keep \textit{private} as a shared
variable---we only did so to be able to show and check the invariants.
\autoref{fig:tas2} shows a more straightforward implementation of a spinlock.
% The solution is similar to the na\"{i}ve solution of \autoref{fig:naivelock},
% but uses test-and-set to check and set the lock variable atomically.
The lock is also cleared in an atomic statement to prevent a data race.
This approach is general for any number of threads.

You can test the spinlock with the program in \autoref{fig:cssynch}
using the command \texttt{harmony -m synch=taslock code/cssynch.hny}.
The \texttt{-m} flag tells Harmony to use the \texttt{taslock.hny} file
for the <{synch}> module rather than the standard <{synch}> module
(which contains only a specification of the lock methods).

\index{starvation}%
The spinlock implementation suffers potentially from \emph{starvation}:
an unlucky thread may never be able to get the lock while other threads
successfully acquire the lock one after another.  It could even happen
with just two threads: one thread might successfully acquire the lock
repeatedly in a loop, while another thread is never lucky enough to
acquire the lock in between.
% \autoref{ch:starvation} will talk more about starvation and how
% to prevent it.

A \emph{ticket lock} (\autoref{fig:ticket} is an implementation of a
lock that prevents starvation using an atomic \emph{fetch-and-increment}
operator.  It is inspired by European bakeries.  A European bakery often has a
clearly displayed counter (usually just two digits) and a ticket dispenser.
Tickets are numbered 0 through 99 and repeat over and over again
(in the case of a two digit counter).
When a customer walks into the bakery, they draw a number from the dispenser
and wait until their number comes up.
Every time a customer has been helped, the counter is incremented.
(Note that this only works if there can be no more than 100 customers in
the bakery at a time.)

\autoref{fig:ticket} similarly uses two variables for a lock,
\textit{counter} and \textit{dispenser}.  When a thread acquires the
lock, it fetches the current dispenser value and increments it
modulo \texttt{MAX\_THREADS}, all in one atomic operation.
In practice, \texttt{MAX\_THREADS} would
be a number like $2^{32}$ or $2^{64}$, but since the Harmony model checker
checks every possible state, limiting \texttt{MAX\_THREADS}
to a small number
significantly reduces the time to model check a Harmony program.
Plus it is easier to check that it fails when you run it with more than
\texttt{MAX\_THREADS} threads.
You can test the implementation using the command
\texttt{harmony -m synch=ticket code/cssynch.hny}.
To see it fail, try
\texttt{harmony -c NTHREADS=10 -m synch=ticket code/cssynch.hny}.

@@

In most operating systems, threads are virtual (as opposed to ``raw CPU cores'')
and can be suspended until some condition changes.
For example, a thread that is trying to acquire a lock can be suspended until the lock is
available.
In Harmony, a thread can suspend itself and save its context (state) in a
shared variable.  Recall that the context of a thread contains
its program counter, stack, and register (containing the current
method's variables).
A context is a regular (if complex) Harmony value.
The syntax of the expression that a thread executes to suspend itself
is as follows:

@@

You will find that using the \emph{implementation} of a lock instead
of the \emph{specification} of a lock (in the <{synch}> module)
often leads to the model checker searching a
significantly larger state space.
Thus it makes sense to model check larger programs in a modular fashion:
model check one module implementation at a time, using
specifications for the other modules.

\section*{Exercises}
\begin{problems}
\item
Run \autoref{fig:incfixed} using (i) <{synch}> and (ii) \texttt{synchS}.
Report how many states were explored by Harmony for each module.
\item \label{ex:xy} \autoref{fig:xy} shows a Harmony program with two variables <{x}>
(initially 0) and <{y}> (initially 100) that can be accessed through methods
\texttt{setX} and \texttt{getXY}.  An application invariant is that \texttt{getXY}
should return a pair that sums to 100.  Add the necessary synchronization code.
\item \label{ex:trylock} Implement \texttt{tryAcquire}(\textit{b}) as an additional
interface for both the <{synch}> and \texttt{synchS} modules.
This interface is like \texttt{acquire}(\textit{b}) but never blocks.  It
returns <{True}> if the lock was available (and now acquired) or <{False}>
if the lock was already acquired.
Hint: you do not have to change the existing code.
\item People who use an ATM often first check their balance and then withdraw
a certain amount of money not exceeding their balance.  A negative balance
is not allowed.  \autoref{fig:atm} shows two operations on bank accounts:
one to check the balance and one to withdraw money.
Note that all operations on accounts are carefully protected by a lock
(i.e., there are no data races).
The \texttt{customer}
method models going to a particular ATM and withdrawing money not exceeding
the balance.
Running the code through Harmony reveals that there is a bug.
It is a common type of concurrency bug known as \emph{Time Of Check Time Of
Execution} (TOCTOE).
\index{Time Of Check Time Of Execution}
\index{TOCTOE}
In this case, by the time the withdraw operation is performed,
the balance can have changed.
Fix the code in \autoref{fig:atm}.  Note, you should leave the
customer code the same. You are only allowed to change the
\texttt{atm\_} methods, and you cannot use the \textbf{atomically}
keyword.
\end{problems}

@@

\begin{figure}
\begin{code}
<{:input:../code/mesa.hny}>
\end{code}
\caption{\harmonylink{modules/synch.hny}
Implementation of condition variables in the <{synch}> module}
\label{fig:cv}
\end{figure}

@@

Harmony provides support for Mesa monitors
through the Harmony <{synch}> module.
\autoref{fig:cv} shows the implementation of
condition variables in the <{synch}> module.
\texttt{Condition}() creates a new condition variable.
It is represented by a dictionary containing
a bag of contexts of threads waiting on the condition variable.
(The \texttt{synchS} library instead uses a list of contexts.)

@@

This completes all the Havender methods.
There is, however, another approach, which is sometimes called deadlock
\emph{avoidance}
\index{deadlock avoidance}%
instead of deadlock \emph{prevention}.
In the case of the Dining Philosophers, we want to avoid the situation where each
diner picks up a fork.  If we can prevent more than four diners from starting to
eat at the same time, then we can avoid the conditions for deadlock from ever
happening.
\autoref{fig:dinersavoid} demonstrates this concept.  It uses a
\emph{counting semaphore} to restrict the number of diners at any time to
four.  A counting semaphore is like a binary semaphore, but can be
acquired a given number of times.  It is supported by the <{synch}>
module.  The \texttt{P} or ``procure'' operation acquires a counting
semaphore.  That is, it tries to decrement the semaphore, blocking while
the semaphore has a value of 0.  The \texttt{V} or ``vacate'' operation
increments the semaphore.

@@

The <{synch}> library supports a synchronized message queue,
similar to the \texttt{Queue} object in Python.
Its interface is as follows:
\begin{itemize}
\item \texttt{Queue()} returns a new message queue;
\item <{put}>(<{q}>, \textit{item}) adds \textit{item} to the queue pointed to by <{q}>;
\item <{get}>(<{q}>) waits for and returns an item on the queue pointed to by <{q}>.
\end{itemize}

@@

It is useful to create an abstract network that reliably sends messages
between threads, much like the FIFO queue in the <{synch}> module.
For this, we need a network protocol that Alice and Bob can run.
In particular, it has to be the case that if Alice sends application
messages $m_1, ..., m_n$ in that order, then if Bob receives an
application message
<{m}>, then $m = m_i$ for some <{i}> and Bob will already have received
application messages $m_1, ..., m_i$ (safety).
Also, if the network is fair and Alice sends application message
<{m}>, then eventually Bob should deliver <{m}> (liveness).

@@

\section{The <{synch}> module}

\index{synch module}%
The <{synch}> module provides the following methods:
